---
title: "TidyTuesday_Dec"
output: html_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
library(usethis)
library(tidytuesdayR)
library(tidyverse)
library(gt)
library(gtsummary)
library(ggplot2)
library(janitor)
#library(sf)
library(BAMMtools)
library(magick)
library(camcorder)
library(gifski)
library(quanteda)
library(tidytext)
library(naniar)
library(wordcloud2)
library(classInt)
library(ggbump)
library(gganimate)

```

## Load Data


```{r getdata, echo=FALSE}
pfdattdyr <- tidytuesdayR::tt_load('2024-12-10') 

pf_dat1 <-pfdattdyr$parfumo_data_clean %>% 
        clean_names()

#After data exploration these data are not really clean. Most of the numbers are not unique identifiers nor do perfumes have them
#Secondly, there are 10 records with repeated columnames

pf_dat <- pf_dat1 %>% 
            filter(brand!='Brand')


```


## Explore missingness

```{r exp1, echo=FALSE}

#full records - but missing numbers
pf_dat %>% 
    filter(if_all(c(release_year, concentration, rating_value, main_accords, top_notes, middle_notes, base_notes), ~!is.na(.)))

#records to throw out due to not having valuable information
pf_dat %>% 
    filter(if_all(c(release_year, concentration, rating_value, main_accords, top_notes, middle_notes, base_notes), ~is.na(.)))

gg_miss_var(pf_dat)

```

# Which brands have the most missing data on parfumo?

```{r missinginfo, echo=FALSE}
pf_dat %>% 
    filter(if_all(c(release_year, concentration, rating_value, main_accords, top_notes, middle_notes, base_notes), ~is.na(.))) %>% 
    group_by(brand) %>% 
    add_count() %>% 
    ungroup() %>% 
    filter(n>10) %>% 
    arrange(desc(n)) %>% 
    select(number, name, brand, url, n)

#The top 25 brands missing data on parfumo
pf_dat %>% 
    filter(if_all(c(release_year, concentration, rating_value, main_accords, top_notes, middle_notes, base_notes), ~is.na(.))) %>% 
    count(brand) %>% 
    arrange(desc(n)) %>% 
    print(n=25)

topmiss <-pf_dat %>% 
    filter(if_all(c(release_year, concentration, rating_value, main_accords, top_notes, middle_notes, base_notes), ~is.na(.))) %>% 
    count(brand) %>% 
    arrange(desc(n)) %>% 
    top_n(25) %>% 
    pull(brand)


```


## Categories of notes
### Floral, herb, spice, fruit(citrus, tree), Spirits(whiskey), wood, 
### 'which includes woody, floral, amber, and fresh'

```{r exp2,echo=FALSE}
pf_dat %>% 
    filter(if_all(c(top_notes, middle_notes, base_notes), ~!is.na(.))) %>% 
    select(release_year, main_accords, top_notes, middle_notes, base_notes) %>% 
    filter(if_all(c(main_accords, release_year), ~!is.na(.)))
# Some reading
## https://odeuropa.eu/objectives-timeline/

```

# what is the timeline of the release years?
```{r timeexp, echo=FALSE}
pf_dat %>% 
    filter(!is.na(main_accords)) %>% 
    count(release_year) %>%
    ggplot(aes(x = release_year, y = n)) +
    ylim(0,1750) +
    geom_line() +
    geom_point() +
    labs(title = "Timeline of Year Counts", x = "Year", y = "Count")

#data goes from 1709 to 2024 but 2024 is not 'complete'

```

# Words and their exploration
```{r olfacttextexp, echo=FALSE}
word_counts_accords <- pf_dat %>%
    filter(!is.na(main_accords)) %>% 
    select(main_accords) %>%
    tidytext::unnest_tokens(word, main_accords) %>%
    count(word, sort = TRUE)

#there are perfumes with notes listed WITHOUT main accords

```

# What are the brand affiliations with different perfumers?

```{r brandperfumer, echo=FALSE}
pf_dat %>% 
    filter(!is.na(perfumers)) %>% 
    select(brand, perfumers) %>% 
    distinct() %>% 
    arrange(brand)


pf_dat %>% 
    filter(!is.na(perfumers)) %>% 
    select(perfumers, brand) %>% 
    group_by(perfumers, brand) %>% 
    distinct() %>% 
    add_count() %>% 
    ungroup() %>% 
    arrange(desc(n)) %>% 
    count(perfumers) %>% 
    arrange(desc(n))

#Alberto Morillas seems to be most prolific in terms of numbers of perfumes on this list and numbers of brands hes worked with, 
#Julien Bedel appears to be second in numbers of perfumes but not in the top 25 for perfumers

#Another way to slice the data is 'Brand owner, Perfume maker' in the Perfumers slot

```

# Making a word cloud of main accords over the years
```{r textmineaccords, echo=FALSE}
#break up the dataset into natural breaks on years
#correctly 
# Assuming you want to plot Jenks breaks for the 'dis' variable
pf_dat2 <- pf_dat %>% 
     filter(!is.na(main_accords) & !is.na(release_year))

#jenks_breaks <- classInt::classIntervals(pf_dat2$release_year, n = 10, style = "jenks")$brks

# Create a new factor variable based on Jenks breaks this is not currently working due to computing
#pf_dat2$year_jenks <- cut(pf_dat2$release_year, breaks = jenks_breaks, include.lowest = TRUE)

pf_dat2$year_brks <- cut(pf_dat2$release_year, breaks = c(1700,1900,1925,1950,1975,1985,1995,2000, 2005, 2010, 2015, 2020, 2025), include.lowest = TRUE)

# Rtutor for the map code
# Create a list of datasets for each model year
release_year_datasets <- pf_dat2 %>%
  group_by(year_brks) %>%
  group_split()



#Function to create dataframes with ranks of the words
create_worddata <- function(data) {
  word_data <- data %>%
    select(main_accords) %>%
    tidytext::unnest_tokens(word, main_accords) %>%
    count(word, sort = TRUE) %>% 
    arrange(desc(n)) %>% 
    mutate(rank = seq_len(nrow(.)))

}

wordcounts <-map(release_year_datasets, create_worddata)

# Function to create a word cloud for each dataset
create_wordcloud <- function(data) {
  word_data <- data %>%
    select(main_accords) %>%
    tidytext::unnest_tokens(word, main_accords) %>%
    count(word, sort = TRUE)
  
  set.seed(100)
  
  wordcloud2(word_data) + WCtheme(1)
}

# Apply the wordcloud function to each dataset
wordclouds <- map(release_year_datasets, create_wordcloud)

# Display the word clouds
wordclouds


```

#Need to detect the differences between datasets
```{r diff, echo=FALSE}


```

#Using a bump chart for the changes, hopefully 20 is not too much
```{r bumptrialdat, echo=FALSE}
years_label <- data.frame(label = levels(pf_dat2$year_brks), value=1:length(levels(pf_dat2$year_brks)))
wordcounts_df <- bind_rows(wordcounts, .id = 'group_id') %>%
                mutate(group_id = as.integer(group_id)) %>% 
                left_join(., years_label, by=c('group_id'='value')) %>% 
                mutate(year_brks = factor(as_factor(label), ordered = TRUE)) %>% #forcing an ordered factor to use max and min in plots
                select(-c(label, group_id)) %>% 
                select(year_brks, everything())

```

```{r bumptrial, echo=FALSE, eval=FALSE}
#This is quite interesting but with 20+ colors, this is not easy to tell what it is
ggplot(wordcounts_df, aes(year_brks, rank, group = word, color=word)) +
    geom_bump() +
    scale_y_reverse()

#let's try labels
ggplot(wordcounts_df, aes(year_brks, rank, group = word, color=word)) +
    geom_bump() +
    scale_y_reverse() +
 geom_text(data = wordcounts_df %>% filter(year_brks == min(year_brks)), 
            aes(label = word), 
            hjust = 1, 
            nudge_x = -0.2) +
  geom_text(data = df %>% filter(year_brks == max(year_brks)), 
            aes(label = word), 
            hjust = 0, 
            nudge_x = 0.2) +
  theme_minimal() +
  theme(legend.position = "none")

#New and improved thanks to rtutor.ai
ggplot(wordcounts_df, aes(year_brks, rank, group = word, color=word)) +
    geom_bump() +
    scale_y_reverse() +
    geom_text(data = wordcounts_df %>% filter(year_brks == min(year_brks)), 
              aes(label = word), 
              hjust = 1, 
              nudge_x = -0.2) +
    geom_text(data = wordcounts_df %>% filter(year_brks == max(year_brks)), 
              aes(label = word), 
              hjust = 0, 
              nudge_x = 0.2) +
    coord_cartesian(xlim = c(0.5, length(levels(wordcounts_df$year_brks)) + 0.5)) +
    theme_minimal() +
    theme(legend.position = "none") 

#trying with manual shapes
ggplot(wordcounts_df, aes(year_brks, rank, group = word, color=word)) +
    geom_bump() +
    scale_y_reverse() +
    geom_point(aes(shape=word)) +
    scale_shape_manual(values=seq(0,22)) +
    geom_text(data = wordcounts_df %>% filter(year_brks == min(year_brks)), 
              aes(label = word), 
              hjust = 1, 
              nudge_x = -0.2) +
    geom_text(data = wordcounts_df %>% filter(year_brks == max(year_brks)), 
              aes(label = word), 
              hjust = 0, 
              nudge_x = 0.2) +
    coord_cartesian(xlim = c(0.5, length(levels(wordcounts_df$year_brks)) + 0.5)) +
    theme_minimal() +
    theme(legend.position = "none") 

```


#could possibly also animate the rankings of the words across the year breaks using gganimate

```{r animatemainaccords, echo=FALSE, eval=FALSE}

'! time data must either be an integer, numeric, POSIXct, Date, difftime, or hms object'


#I think to get geom_bump to work with gganimate need to convert to a time period and then use transition events or transition time

bumpp <-ggplot(wordcounts_df, aes(year_brks, rank, group = word, color=word)) +
    geom_bump() +
    scale_y_reverse() +
    geom_point(aes(shape=word)) +
    scale_shape_manual(values=seq(0,22)) +
    geom_text(data = wordcounts_df %>% filter(year_brks == min(year_brks)), 
              aes(label = word), 
              hjust = 1, 
              nudge_x = -0.2) +
    geom_text(data = wordcounts_df %>% filter(year_brks == max(year_brks)), 
              aes(label = word), 
              hjust = 0, 
              nudge_x = 0.2) +
    coord_cartesian(xlim = c(0.5, length(levels(wordcounts_df$year_brks)) + 0.5)) +
    theme_minimal() +
    theme(legend.position = "none") +
    transition_reveal(year_brks, transition_length = 2, state_length = 1, wrap=FALSE)
    enter_fade() +
    exit_fade()

# Render the animation
animate(bumpp, nframes = 100, fps = 10)


```

#Building the animation from scratch
```{r animationscratch, echo=FALSE}

brks <- levels(wordcounts_df$year_brks)

#takes a list from a vector and starts there
create_dfs <- function(data, value) {
  
    df <-data %>%
     mutate(word = ifelse(year_brks>value, NA, word),
            rank = ifelse(year_brks>value, NA, rank)) 
    return(df)

  }

vec_int <- brks[2:length(brks)] # taking the second value of the vector and running the datasets to the end
seq_dfs <-map(vec_int, ~create_dfs(wordcounts_df, .x))




makeplots <- function(df) {
  ggplot(df, aes(year_brks, rank, group = word, color=word)) +
    geom_bump() +
    scale_y_reverse(breaks = seq(1,21, by=1)) +
    geom_point(aes(shape=word)) +
    scale_shape_manual(values=seq(0,22)) +
    geom_text(data = df %>% filter(year_brks == min(year_brks)), 
              aes(label = word), 
              hjust = 1, 
              nudge_x = -0.2) +
    geom_text(data = df %>% filter(!is.na(word)) %>% filter(year_brks==max(year_brks)), 
              aes(label = word), 
              hjust = 0, 
              nudge_x = 0.2) +
    coord_cartesian(xlim = c(0.5, length(levels(wordcounts_df$year_brks)) + 0.5)) + #not the best function line it calls external dataset
    theme_minimal() +
    theme(legend.position = "none") +
    labs(
    title = "Rank of Main Accords for Parfumes by Release Year",
    subtitle = "Often Floral and Spicy, recently, Sweet",
    caption = "Each parfume may have multiple main accords noted \n Data source:Parfumo via https://www.kaggle.com/olgagmiufana1",
    x = 'Year of Release (binned)'
  )
  
}

plotspack <- map(seq_dfs, makeplots)

```

```{r makegif, echo=FALSE}
gg_record(
    dir = file.path("/cloud/project", 'recording2'),
    device = 'png',
    width = 12, 
    height = 6
)

plotspack

gg_stop_recording()

png_files <- list.files("/cloud/project/recording2", pattern = ".*png$", full.names = TRUE)
gifski(png_files, gif_file = "pf_plot.gif", width = 1200, height = 600, delay = 2)



```
